Investigating Neolithic pottery usage in the Yanyuan Basin in Southwest China using organic residue analysis

### 2023-01-08 analysis/paper/paper.Rmd
worked on the result section

This study aims to detect and identify organic residues, particularly lipid residues, that could have been preserved in the ceramic fabric to explore the vessel usage at Guijiabao. Lipid analysis have proven useful in archaeology since lipid molecules can be well preserved over thousand years in the porous fabric of pottery with the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. Lipid residues are the most direct evidence to investigate culinary practices and dietary patterns since specific lipid compounds can be associated with different types of foodstuffs cooked or stored in pottery vessels [@Copley2005]. It is especially informative for human subsistence in the situation where archaeological sites lack organic remains due to acidic sediments [@Kwak2020]. The growing application of organic residue analysis demonstrate it offers a better understanding of culinary practices during the transitional period to agricultural societies [@Craig2011], dietary patterns in

@@

@@

```{r tidy data}
library(tidyverse)

@@

Lipid residues from our samples show that the saturated fatty acids with even number of carbons (C14:0, C16:0, and C18:0), particularly palmitic (C16:0) and stearic(C18:0), are the dominant compounds. Because palmitic and stearic fatty acids are normally degraded from animal fat or plant wax, the domination pattern with little concentration of other fatty acids suggests the organic compounds in our samples were highly degraded during post depositional processes. However, there is one sample contains rich compounds that include long-chain fatty acids, C20, C22, and C24, which were absent in other samples. This might indicate

To further distinguish the source of fatty acids, we conducted stable carbon isotope analysis for those samples. Here you can see the stable carbon isotope measurements of C16:0 and C18:0 fatty acids. The results suggest a dominance of C3 plant for most pots, and a few shows a mixture of animal fats and C3 plants. The sample 5 where more compounds were preserved shows a source of C4 plant or marine animals instead. We think this is very interesting finding since there was just a few rice grains found at this site compared to the evidence of millet.

@@

# Conclusion
Our results show a mixture of C3 plants and terrestrial animals for most pots with one pot showing more C4 plant sources. This case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region. Our case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.

>>>>

This study aims to detect and identify organic residues, particularly lipid residues, that could have been preserved in the ceramic fabric to explore the vessel usage at Guijiabao. Lipid analysis is useful in archaeology since lipid molecules can be well preserved over thousand years in the porous fabric of pottery with the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017]. Lipid residues are the most direct evidence to investigate culinary practices and dietary patterns since specific lipid compounds can be associated with different types of foodstuffs cooked or stored in pottery vessels [@Copley2005]. It is especially informative for human subsistence in the situation where archaeological sites lack organic remains due to acidic sediments [@Kwak2020]. The growing application of organic residue analysis demonstrate it offers a better understanding of culinary practices during the transitional period to agricultural societies [@Craig2011], dietary patterns in

@@

To further distinguish the source of fatty acids, we conducted stable carbon isotope analysis for those samples.

@@

```{r tidy-data, cache=TRUE, cache.extra=tools::md5sum(here::here("analysis/data/raw_data/LY_21_08_20_pot_chrom.CSV"))}
library(tidyverse)

@@

```{r highlight-compounds, cache=TRUE}
# highlight target compounds
highlight_GJB_four <-
  meth_all %>%
  mutate(retention = round(Path, 1)) %>%
  group_by(sample, retention) %>%
  arrange(desc(File)) %>%
  distinct(retention, .keep_all = TRUE) %>%
  mutate(FA = case_when(
    retention == 21.6 ~ "C14:0",
    retention == 23.6 ~ "C15:0",
    retention == 25.5 ~ "C16:0",
    retention == 27.3 ~ "C17:0",
    retention == 28.4 ~ "C18:1",
    retention == 29.1 ~ "C18:0",
    retention == 32.4 ~ "C20:0",
    retention == 35.4 ~ "C22:0")) %>%
  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3")) %>%
  filter(File > 100000)

highlight_IS <-
  meth_all %>%
  mutate(retention = round(Path, 1)) %>%
  group_by(sample, retention) %>%
  arrange(desc(File)) %>%
  distinct(retention, .keep_all = TRUE) %>%
  mutate(FA = case_when(
    retention == 33.9 ~ "IS",
    retention == 41.3 ~ "IS")) %>%
  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3"))

# highlight the compounds in the potsherd with high concentration
highlight_SYG2_2 <-
  meth_all %>%
  mutate(retention = round(Path, 1)) %>%
  group_by(sample, retention) %>%
  arrange(desc(File)) %>%
  distinct(retention, .keep_all = TRUE) %>%
  mutate(FA = case_when(
    retention == 21.6 ~ "C14:0",
    retention == 25.5 ~ "C16:0",
    retention == 27.3 ~ "C17:0",
    retention == 28.4 ~ "C18:1",
    retention == 29.1 ~ "C18:0",
    retention == 31.0 ~ "C19:0",
    retention == 32.4 ~ "C20:0",
    retention == 35.4 ~ "C22:0",
    retention == 36.8 ~ "C23:0",
    retention == 38.2 ~ "C24:0",
    retention == 40.9 ~ "C26:0",
    retention == 33.9 ~ "IS",
    retention == 41.3 ~ "IS",
    retention == 45.6 ~ "C:30",
    retention == 47.8 ~ "C:32")) %>%
  filter(name == "SYG-TN13-E22-2#2")
```

```{r chromatogram-four, cache=TRUE}
# plot the first four
chroma_four <-
  meth_all %>%
  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3")) %>%
  mutate(File = ifelse(File > 550000, 550000, File)) %>%
  ggplot(aes(Path, File)) +
  geom_line(size = 0.3) +
  geom_text(data = highlight_GJB_four, #ggrepel::geom_text_repel
            aes(label = FA, angle = 90),
            size = 3,
            nudge_y = 87000,
            show.legend = FALSE) +
  geom_text(data = highlight_IS,
            aes(label = FA),
            size = 3,
            nudge_y = 40000,
            show.legend = FALSE) +
  geom_segment(aes(x= 41.1, xend= 41.5, y= 550000, yend= 550000), size =0.3) +
  geom_segment(aes(x= 41.1, xend= 41.5, y= 560000, yend= 560000), size =0.3) +
  annotate("text", x = 41.3, y = 600000, label = "IS", size =3) +
  facet_wrap(~name, ncol = 1) +
  scale_y_continuous(labels = scales::comma_format(),
                     limits = c(0, 600000),
                     breaks = seq(0, 600000, 200000)) +
  scale_x_continuous(limits = c(20, 50),
                     breaks = seq(20, 50, 5),
                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
  labs(x = "retention time", y = "relative intensity") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

ggsave(plot = chroma_four,
       here::here("analysis","figures", "chromatograms_four.png"),
       width = 7,
       height = 8,
       dpi = 300,
       units = "in")
```

The GC-MS analysis for organic residues extracted from Guijiabao potsherds revealed the presence of lipid compounds, but with low amounts in most of them. The lipid profiles show that the saturated fatty acids with an even number of carbons, particularly C16:0 (palmitic acid ) and C18:0 (stearic acid), and monounsaturated fatty acid  C18:1 (oleic acid), are the main compounds for most potsherds (Figure \@ref(fig:chromatograms-plots-four)). Those fatty acids commonly exist in most animal fat and plant oil. They are also generally degraded from other larger molecules such as triacylglycerols due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of C16:0 and C18:0 fatty acids in our samples suggest that the original organic residues were highly degraded during post-depositional processes. In addition, most potsherds present higher contents of C18:1 compared to C18:0 and C16:0, such as SYG-TN13-E23-3. A larger content of C18:1 and C16:0 with smaller content of C18:0 could indicate an origin of plant oils [@Mayyas2018; @Copley2005]. Although oleic acid (C18:1) has the highest concentration in most of our potsherd samples, the generally low concentrations of all fatty acids with minor differences in their amounts may not fully support a plant origin.

```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from Guijiabao. Cn:x refers to fatty acid with n carbon atoms and x double bonds. IS indicates internal standard used for quantification, including FA C21:0 and n-alkane 34"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
```

```{r chromatogram-specific, cache=TRUE}
# plot the specific one
chroma_specific <-
  meth_all %>%
  filter(name == "SYG-TN13-E22-2#2") %>%
  ggplot(aes(Path, File)) +
  geom_line(size = 0.3) +
  geom_text(data = highlight_SYG2_2, #ggrepel::geom_text_repel
            aes(label = FA),
            size = 3,
            nudge_y = 250000,
            show.legend = FALSE) +
  scale_y_continuous(labels = scales::comma_format(),
                     limits = c(0, 7500000),
                     breaks = seq(0, 7500000, 1000000)) +
  scale_x_continuous(limits = c(15, 50),
                     breaks = seq(15, 50, 5),
                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
  labs(title = "SYG-TN13-E22-2#2", x = "retention time", y = "relative Intensity") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

ggsave(plot = chroma_specific,
       here::here("analysis","figures", "chromatograms_specific.png"),
       width = 8,
       height = 5,
       dpi = 300,
       units = "in")
```

Long-chain fatty acids from C20:0 to C32:0 are present in only one pottery sample (SYG-TN13-E22-2#2) with an even-over-odd carbon chain length predominance, but in lower abundance than the medium-chain components (Figure \@ref(fig:chromatograms-specific)). The existence of long-chain fatty acids with a high abundance of C16:0 relative to C18:0 might be related to plant waxes [@Post1996; @Copley2005]. However, it could be also associated with an origin of degraded animal fat based on a strong predominance of medium-chain fatty acids [@Whelton2018; @Dudd1998]. Overall, the GC-MS results demonstrate evidence of degraded animal fats or plant oil.

```{r chromatograms-specific, echo=FALSE, fig.cap= "GC-MS chromatograms of the potshard with the presence of long-chain fatty acids. Cn:x refers to fatty acid with n carbon atoms and x double bonds. IS indicates internal standard used for quantification, including FA C21:0 and n-alkane 34"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_specific.png"))
```

```{r isotopic-analysis}
library(tidyverse)
isotope_data <-
  readxl::read_excel(here::here("analysis","data", "raw_data","12_21_potshard_13C.xlsx"),
                     sheet = "Summary", col_names =  FALSE)

# filter the Meth corrected data
delta_meth_cor <-
  isotope_data %>%
  select(1, 7, 9) %>%
  slice(2:8) %>%
  dplyr::mutate(...1 = ifelse(is.na(...1), paste0("C13"), ...1))

names(delta_meth_cor) <- delta_meth_cor[1,]

delta_meth_cor <-
  delta_meth_cor %>%
  slice(2:7) %>%
  dplyr::mutate(across(starts_with("13C"), as.numeric)) %>%
  dplyr::mutate(delta = `13C C18:0`-`13C C16:0`) %>%
  mutate(sample = case_when(
    C13 %in% c("Pot 1") ~ "CDG-062",
    C13 %in% c("Pot 2") ~ "SYG-3",
    C13 %in% c("Pot 3") ~ "SYG-TN13-E22-2#1",
    C13 %in% c("Pot 4") ~ "SYG-TN13-E22-1",
    C13 %in% c("Pot 5") ~ "SYG-TN13-E22-2#2",
    C13 %in% c("Pot 6") ~ "SYG-TN13-E23-3")) %>%
  mutate(sample = factor(sample, levels = c("SYG-TN13-E22-1",
                                            "SYG-TN13-E22-2#1",
                                            "SYG-TN13-E22-2#2",
                                            "SYG-TN13-E23-3",
                                            "SYG-3",
                                            "CDG-062")))

delta_min <- round(min(delta_meth_cor$delta), 2)
delta_max <- round(max(delta_meth_cor$delta), 2)

# plot the carbon isotopes 
isotope_C16_C18 <-
  ggplot(delta_meth_cor, #delta_meth_cor/delta_blank_cor
       aes(`13C C16:0`,`13C C18:0`)) +
  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
  #ggrepel::geom_text_repel(aes(label = C13)) +
  theme_minimal() +
  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
       y = bquote(delta*{}^13*"C 18:0 \u2030")) +
  xlim(-33,-20) +
  ylim(-33,-20) +
  #coord_fixed(ratio = 1) +
  geom_abline(intercept = -0.7, linetype = "dashed") + #Suryanarayan et al.2021
  geom_abline(intercept = -3.1, linetype = "dashed") +
  annotate("text", x = -21.5, y = -22, angle = 45, vjust = 1.5,
           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
  annotate("text", x = -21.5, y = -24.5, angle = 45, vjust = 1.5,
           label = bquote(Delta*{}^13*"C = -3.1 \u2030"))

big_delta_C16 <-
  ggplot(delta_meth_cor, # delta_meth_cor or delta_blank_cor
       aes(`13C C16:0`, delta)) +
  geom_point(size = 2, alpha = 0.9, aes(shape = sample)) +
  theme_minimal() +
  labs(x = bquote(delta*{}^13*"C 16:0 \u2030"),
       y = bquote(Delta*{}^13*"C 18:0 \u2030")) +
  scale_x_continuous(limits = c(-33, -20),
                     breaks = seq(-33, -20, 2)) +
  ylim(-2.5, 2) +
  geom_hline(yintercept = -0.7, linetype = "dashed") +
  annotate("text", x = -22, y = -0.5,
           label = bquote(Delta*{}^13*"C = -0.7 \u2030")) +
  annotate("text", x = -31, y = 1.8,
           label = "C3-diet") +
  annotate("segment", x = -29, y = 1.8, xend = -24, yend = 1.8,
           arrow = arrow(type = "closed", length = unit(0.02, "npc"))) +
  annotate("text", x = -22, y = 1.8,
           label = "C4-diet")

library(cowplot)
combined_isotopic_plots <-
  plot_grid(isotope_C16_C18,
          big_delta_C16,
          labels = c('A', 'B'),
          ncol =1,
          align = 'v',
          label_size = 12)

ggsave(here::here("analysis","figures","isotopic_plots.png"),
       width = 6,
       height = 8,
       dpi = 300,
       units = "in")
```

To identify the origins of degraded animal fats, we conducted isotope analysis to obtain the δ^13^C values of C16:0 and C18:0 using GC-C-IRMS. The δ^13^C values of C16:0 and C18:0 have proven useful to distinguish between ruminant adipose, non-ruminant adipose, and ruminant dairy fats based on different ways where dietary carbon sources are utilized by different fat types [@Dunne2012; @Copley2003; @Evershed2008]. To reduce the bias caused by differential δ^13^C values across geographical regions, we used the Δ^13^C(δ^13^C18:0 - δ^13^C16:0) values to determine the origins of degraded fats [@Evershed2008; @Salque2013]. The isotopic results show the Δ13C values in our samples range from `r delta_min` ‰ to `r delta_max` ‰ that were plotted according to the global reference database for the identification of their origins [@Suryanarayan2021; @Dunne2012; @Evershed2008].

Most of the potsherd samples from Guijiabao contain fatty acids from ruminant adipose fats with one potsherd (SYG-TN13-E23-3) suggesting an origin of non-ruminant adipose fat (Figure \@ref(fig:isotopic-plots): A). The only potsherd from Gaushan (CDG-062) demonstrates ruminant adipose fats as well. There are two potsherds (SYG-TN13-E22-1 and SYG-3) from Guijiabao plotted between the reference line of ruminant adipose and non-ruminant, indicative of the mixing of those two sources [@Copley2003]. Moreover, the potsherd indicating non-ruminant adipose fats has a relatively depleted δ^13^C value (Figure \@ref(fig:isotopic-plots): B) that falls within the range of C3/terrestrial plants [@Shoda2018; @Kwak2020]. In general, most potsherd samples containing ruminant adipose fats have depleted δ^13^C values consistent with C3/terrestrial plants, suggesting the ruminant animals were likely fed on C3 plants. There is only one potsherd (SYG-TN13-E22-2#2) have a relatively enriched δ^13^C value, indicative of a C4 plant diet for the ruminant animal.

```{r isotopic-plots, echo=FALSE, fig.cap= ""}
knitr::include_graphics(here::here("analysis/figures/isotopic_plots.png"))
```

@@

# Conclusion
Our investigation on the ceramics excavated from Southwest China suggests that a mixture of C3 plants and ruminant animals could be the food sources in this region during the Neolithic transition. This case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region. Our case study not only indicates the possibility of rice agriculture that has not been identified previously at the site but also demonstrates the first use of lipid analysis for pottery to explore the diet of the late Neolithic period in this region.
